\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2017-01-24 doc/publications/plos16/plos16.tex
Fix figure and table refs

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This section talks about the datasets
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Datasets}
\subsubsection{1000 Genomes Project}
We obtained the 1000 Genomes Project data as VCF files from their FTP site.  Each VCF file contains the variants for
every individual across one chromosome.  We use the phase 3 dataset which contains 2,504 individuals with approximately
81 million variant positions (i.e. a matrix of $2,504 \times 81$ million).  For compatibility with the majority of the
algorithms we demonstrate, we convert the original variant representations to double representations.  For example, we
store the absence of a variant (0\textbar0) as 0, a heterozygous
variant (1\textbar0 or 0\textbar1) as 1, and a
homozygous variant (1\textbar1) as 2.  This is not the only possible encoding.  \cite{Goldstein.et.al.2011} note that of
the encoding listed in table \ref{table:encodings}, the additive encoding will present the most difficulty for a random
forest model.

\begin{table}[!ht]
%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\begin{flushleft} 
\centering
\caption{\bf Possible encodings of GWAS data \cite[]{Goldstein.et.al.2011}}
\begin{tabular}{|l|l|l|l||}
\hline
{\bf Type} & {\bf Mechanism}                                         & {\bf Partition} \\ 
\thickhline
Additive   & Each additional minor allele increases variation        & 0, 1, 2         \\
Dominant   & Presence of at least 1 minor allele increases variation & 0, 1/2          \\
Recessive  & Two minor alleles needed for variation                  & 0/1, 2          \\
Heterosis  & Heterozygote leads to variation                         & 0/2, 1          \\ \hline
%\end{adjustwidth}
\end{tabular}
\label{table:encodings}
\end{flushleft}
\end{table}

@@

% Fig.~\ref{figure:out-of-bag-prediction-error-prod.png} illustrates that increasing the number of predictors sampled for splitting at each node
% (\mtry) increases the accuracy with which features are ranked.  From $mtry=0.5$ the sort order of the features reflects
% the weight originally assigned to them (dotted line). Further increasing \mtry\ improves the predicted importance
% measure, which starts to converge to the weight originally assigned to them (coefficients).  It also increases the
% difference between the associated 5 variable and the remaining variables in the top 20 list.  By $mtry=0.8$ the variable
% with the smallest weight has a predicted importance measure that is 5 times larger than the aggregated importance of the
% 15 remaining variables from the list.

% Fig.~\ref{figure:rbo-prod.png} illustrates that increasing \mtry\ also improves the OOB
% (out of bag) classification error.
% By $mtry=0.8$ the error is converging to X, which is due to Y.
% \marginpar{what are ``X'' and ``Y''. Piotr}

@@

\marginpar{this was done with n=10000. Needs to redone with n=2.5M and some things re-written}
Fig.~\ref{figure:synth1} illustrates the effect of varying \mtry and \ntree on these two measures. The default \mtry
value for a classification problem is $\sqrt{p}=100$ so in this case we are considering values considerably larger than
the default. Fig.~\ref{figure:out-of-bag-prediction-error-prod.png} shows the OOB error is minimized at the largerst
values of \mtry and \ntree considered  whereas Fig.~\ref{figure:rbo-prod.png} illustrates that the RBO measure is less
sensitive once we over a certain threshold.

@@

\cursedforest solves this problem of high-dimensionality genomic data. Not only is it faster on the above smaller datasets,
but it is also able to scale to the entire genome. In table~\ref{cursedforesttable} we present our observed times building
models with CursedForest using all available resources on our Spark cluster.

@@

First we look at \cursedforest horizontal scalability for a medium size dataset of 2.5 million variables and 5000
samples, by varying the \mtry  fraction and the number of CPU cores allocated to the execution. 
The results are  visualized in figure \ref{figure:synthetictiming} below. 
%The raw results are
%presented in table \ref{synthetictimingtable} and visualized in figure \ref{figure:synthetictiming} below. 
Regardless of the number of cores used, \cursedforest displays approximately linear dependency between the execution
time and \mtry (Fig. \ref{figure:synthetictiming.a}).

\cursedforest scales almost linearly with the number of CPU cores for medium values of \mtry fraction but for both
lower and higher values the performance degrades slightly (Fig \ref{figure:synthetictiming.b}). In the latter case the likely cause is
communication overhead (with lower \mtry values the proportion of time for parallelizable computation to the time for
internode communication is lower) while in the latter case it is most likely caused by reaching the clusters
computational capacity.

\marginpar{tables 2 has been commented out -- there is no data in it. Piotr could add the data or we could remove both
  tables just leave the figures. I have done this for the time being} 
Next we investigate \cursedforest scalability with
regards to the size of data, by varying the number of variables and sample for a fixed \mtry fraction of 0.25 and
execution of 128 CPU cores. The results are visualized in figure \ref{figure:synth2} below (please note log scale on the
axes and the values on y axes are expresses as trees per hour).

@@

Donoho and Tanner \cite{Donoho.and.Tanner.2009} give a ``universal phase change'' result that has applications in a
large number of areas including variable selection in high dimensions. Consider figure
\ref{figure:phase-diagram-equivalence.png} which shows the region where a model can recover the important variables,
plotted as a function of $\delta = n/p$ and $\rho =k/n$ (where $k$ is the number of significant variables). There is a
distinct boundary, shown empirically and also based on arguments from combinatorial geometry, to the region where we can
reliably recover significant variables.  \cite{Donoho.and.Stodden.2006} investigate the behaviour of a number of
regression approaches for variable selection (LARS, Lasso and forward stepwise) and make the point that above the
phase-transition line variable recovery is still possible by a combinatorial approach.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This section talks about the datasets
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Datasets}
\subsubsection{1000 Genomes Project}
We obtained the 1000 Genomes Project data as VCF files from their FTP site.  Each VCF file contains the variants for
every individual across one chromosome.  We use the phase 3 dataset which contains 2,504 individuals with approximately
81 million variant positions (i.e. a matrix of $2,504 \times 81$ million).  For compatibility with the majority of the
algorithms we demonstrate, we convert the original variant representations to double representations.  For example, we
store the absence of a variant (0\textbar0) as 0, a heterozygous
variant (1\textbar0 or 0\textbar1) as 1, and a
homozygous variant (1\textbar1) as 2.  This is not the only possible encoding.  \cite{Goldstein.et.al.2011} note that of
the encoding listed in Table~\ref{table:encodings}, the additive encoding will present the most difficulty for a random
forest model.

\begin{table}[!ht]
%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\begin{flushleft} 
\centering
\caption{\bf Possible encodings of GWAS data \cite{Goldstein.et.al.2011}}
\begin{tabular}{|l|l|l|l||}
\hline
{\bf Type} & {\bf Mechanism}                                         & {\bf Partition} \\ 
\thickhline
Additive   & Each additional minor allele increases variation        & 0, 1, 2         \\
Dominant   & Presence of at least 1 minor allele increases variation & 0, 1/2          \\
Recessive  & Two minor alleles needed for variation                  & 0/1, 2          \\
Heterosis  & Heterozygote leads to variation                         & 0/2, 1          \\ \hline
%\end{adjustwidth}
\end{tabular}
\label{table:encodings}
\end{flushleft}
\end{table}

@@

% Fig~\ref{figure:out-of-bag-prediction-error-prod.png} illustrates that increasing the number of predictors sampled for splitting at each node
% (\mtry) increases the accuracy with which features are ranked.  From $mtry=0.5$ the sort order of the features reflects
% the weight originally assigned to them (dotted line). Further increasing \mtry\ improves the predicted importance
% measure, which starts to converge to the weight originally assigned to them (coefficients).  It also increases the
% difference between the associated 5 variable and the remaining variables in the top 20 list.  By $mtry=0.8$ the variable
% with the smallest weight has a predicted importance measure that is 5 times larger than the aggregated importance of the
% 15 remaining variables from the list.

% Fig~\ref{figure:rbo-prod.png} illustrates that increasing \mtry\ also improves the OOB
% (out of bag) classification error.
% By $mtry=0.8$ the error is converging to X, which is due to Y.
% \marginpar{what are ``X'' and ``Y''. Piotr}

@@

\marginpar{this was done with n=10000. Needs to redone with n=2.5M and some things re-written}
Fig~\ref{figure:synth1} illustrates the effect of varying \mtry and \ntree on these two measures. The default \mtry
value for a classification problem is $\sqrt{p}=100$ so in this case we are considering values considerably larger than
the default. Fig~\ref{figure:out-of-bag-prediction-error-prod.png} shows the OOB error is minimized at the largerst
values of \mtry and \ntree considered  whereas Fig~\ref{figure:rbo-prod.png} illustrates that the RBO measure is less
sensitive once we over a certain threshold.

@@

\cursedforest solves this problem of high-dimensionality genomic data. Not only is it faster on the above smaller datasets,
but it is also able to scale to the entire genome. In Table~\ref{cursedforesttable} we present our observed times building
models with CursedForest using all available resources on our Spark cluster.

@@

First we look at \cursedforest horizontal scalability for a medium size dataset of 2.5 million variables and 5000
samples, by varying the \mtry  fraction and the number of CPU cores allocated to the execution. 
The results are  visualized in Fig~\ref{figure:synthetictiming} below. 
%The raw results are
%presented in Table~\ref{synthetictimingtable} and visualized in figure \ref{figure:synthetictiming} below. 
Regardless of the number of cores used, \cursedforest displays approximately linear dependency between the execution
time and \mtry (Fig~\ref{figure:synthetictiming.a}).

\cursedforest scales almost linearly with the number of CPU cores for medium values of \mtry fraction but for both
lower and higher values the performance degrades slightly (Fig~\ref{figure:synthetictiming.b}). In the latter case the likely cause is
communication overhead (with lower \mtry values the proportion of time for parallelizable computation to the time for
internode communication is lower) while in the latter case it is most likely caused by reaching the clusters
computational capacity.

\marginpar{tables 2 has been commented out -- there is no data in it. Piotr could add the data or we could remove both
  tables just leave the figures. I have done this for the time being} 
Next we investigate \cursedforest scalability with
regards to the size of data, by varying the number of variables and sample for a fixed \mtry fraction of 0.25 and
execution of 128 CPU cores. The results are visualized in Fig~\ref{figure:synth2} below (please note log scale on the
axes and the values on y axes are expresses as trees per hour).

@@

Donoho and Tanner \cite{Donoho.and.Tanner.2009} give a ``universal phase change'' result that has applications in a
large number of areas including variable selection in high dimensions. Consider
Fig~\ref{figure:phase-diagram-equivalence.png} which shows the region where a model can recover the important variables,
plotted as a function of $\delta = n/p$ and $\rho =k/n$ (where $k$ is the number of significant variables). There is a
distinct boundary, shown empirically and also based on arguments from combinatorial geometry, to the region where we can
reliably recover significant variables.  \cite{Donoho.and.Stodden.2006} investigate the behaviour of a number of
regression approaches for variable selection (LARS, Lasso and forward stepwise) and make the point that above the
phase-transition line variable recovery is still possible by a combinatorial approach.
